A Quasi-Genuine FIFO Total Order Multicast Primitive

### 2011-06-14 textos/optpaxos/draftv3.tex
it may be possible to deliver conservatively in 3delta EVEN in the worst case

\begin{itemize}
  \item the maximum message delay bound $\delta$ is known;
  \item at some point in the execution of $I$, there is a leader which does not fail until a value has been chosen\footnote{Lamport demonstrates in \cite{lamport1998ptp} that, if the time needed to elect a leader is $T_{el}$, the time needed to conclude a consensus instance would be at most $T_{el} + 9\delta$ after the last leader failure during the execution of $I$.};
  \item during the whole execution of $I$, more than half of the processes are correct, that is, if the number of faulty processes is $f$ and the total number of processes is $n$, $f < \lceil \frac{n}{2} \rceil$;
  \item enough messages are successfully received, that is, for each phase of the execution of $I$, the leader process successfully receives the reply (be it a confirmation or a negative acknowledgement) from a majority of processes.
\end{itemize}

@@

In the context of our \cms{} primitive, the first communication step of Paxos -- forwarding a message $m$ to the leader -- is already done by means of \rmc{}ing each message to a set of processes that includes every process in $m$'s group of origin. The last phase, which consists of every learner receiving the confirmation of a majority of acceptors, can also include the $\delta$ included in the conservative message delivery time we analysed before: instead of waiting for a message to be decided and only then notifying other groups, the acceptors of a group can also send the confirmation to processes in other groups, so that also they can infer the first group's decision right away. This implies that, for each consensus instance run within a group $G$, each process in the groups of $receiversFrom(G)$ would be a learner for that instance. Figure \ref{fig:paxosmanylearnergroups} illustrates this: process $p_1$ from group $G$, whose leader is $p_3$, \cms{}s $m$; in the last phase of the consensus protocol, the processes in $G' \in receiversFrom(G)$ receive the confirmation message from the acceptors of $G$, so no extra communication step is needed to notify them about the final timestamp of $m$.

As $m$ was \rmc{}ed in the beginning, there is no need to send it again in the consensus instance. Instead, only the timestamp of $m$ is proposed, since this is the information that must be agreed upon -- and which may change if the optimistic assumption does not hold and some other message of higher timestamp is delivered before $m$. In any case, as all messages agreed upon in $G$ are notified to $G'$, also the processes of $G'$ are able to infer the final timestamp of each message from $G$.

@@

Note, however, that for this to happen it would be necessary for the leader $p$ of group $G$ to: (1) not know some value greater than or equal to $w(p)$; (2) have received and proposed some message $m' : m'.ts > m.ts$ before $m$ and (3) not have received any barrier with value greater than the new timestamp of $m$. Even after changing the timestamp of $m$, it may be not necessary to request a new barrier from each group in $sendersTo(G)$, because some other barrier -- possibly requested because of some other message -- with a value higher than the new timestamp of $m$ may have been received already.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/paxos3d}
  \caption{Deciding the final timestamp and notifying other groups in $3\delta$}
  \label{fig:paxosmanylearnergroups}
\end{figure}

@@

Even if it is unlikely to happen, the worst case delivery time of \cms{} when using Paxos with a leader for consensus and requesting barriers in the beginning of the protocol is $6\delta$. It is so because some messages may need to have their initial timestamp changed as a possible consequence of the optimistic assumption \emph{A2} not holding. However, a simple way to eliminate the possibility of having to change timestamps is by not considering messages whose delivery order does not follow the order of the initial timestamps.

@@

The problem is that this would violate the uniform validity property, which could make such multicast primitive impracticable for some applications. Nevertheless, some other applications not only do not require messages to be delivered if they arrive out of order, but also it is better not to deliver such messages at all -- such as real-time streams of audio or video or online real-time games.

>>>>

\begin{itemize}
  \item the maximum message delay bound $\delta$ is known;
  \item at some point in the execution of $I$, there is a leader which does not fail until a value has been chosen\footnote{Lamport demonstrates in \cite{lamport1998ptp} that, if the time needed to elect a leader is $T_{el}$, the time needed to conclude a consensus instance would be at most $T_{el} + 9\delta$ after the last leader failure during the execution of $I$.};
  \item during the whole execution of $I$, more than half of the processes are correct, that is, if the number of faulty processes is $f$ and the total number of processes is $n$, $f < \lceil \frac{n}{2} \rceil$;
  \item enough messages are successfully received, that is, for each phase of the execution of $I$, a majority of processes succesfully receive the leader's message or the leader successfully receives the reply (be it a confirmation or a negative acknowledgement) from a majority of processes.
\end{itemize}

@@

In the context of our \cms{} primitive, the first communication step of Paxos -- forwarding a message $m$ to the leader -- is already done by means of \rmc{}ing each message to a set of processes that includes every process in $m$'s group of origin. The last phase, which consists of every learner receiving the confirmation of a majority of acceptors, can also include the $\delta$ contained in the conservative message delivery time we analysed before: instead of waiting for a message to be decided and only then notifying other groups, the acceptors of a group can also send the confirmation to processes in other groups, so that also they can infer the first group's decision right away. This implies that, for each consensus instance run within a group $G$, each process in the groups of $receiversFrom(G)$ would be a learner for that instance. Figure \ref{fig:paxosmanylearnergroups} illustrates this: process $p_1$ from group $G$, whose leader is $p_3$, \cms{}s $m$; in the last phase of the consensus protocol, the processes in $G' \in receiversFrom(G)$ receive the confirmation message from the acceptors of $G$, so no extra communication step is needed to notify them about the final timestamp of $m$.

As $m$ was \rmc{} in the beginning, there is no need to send all of its contents again in the consensus instance. Instead, only the timestamp of $m$ is proposed, since this is the information that must be agreed upon -- and which may change if the optimistic assumption does not hold and some other message of higher timestamp is delivered before $m$. In any case, as all messages agreed upon in $G$ are notified to $G'$, also the processes of $G'$ are able to infer the final timestamp of each message from $G$.

@@

Note, however, that for this worst case scenario to happen it would be necessary for the leader $p$ of group $G$ to: (1) not know some value greater than or equal to $w(p)$; (2) have received and proposed some message $m' : m'.ts > m.ts$ before $m$ and (3) not have received any barrier with value greater than the new timestamp of $m$. Even after changing the timestamp of $m$, it may be not necessary to request a new barrier from each group in $sendersTo(G)$, because some other barrier -- possibly requested because of some other message -- with a value higher than the new timestamp of $m$ may have been received already.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/paxos3d}
  \caption{Deciding the final timestamp and notifying other groups in three communication steps}
  \label{fig:paxosmanylearnergroups}
\end{figure}

@@

Although unlikely to happen, the worst case delivery time of \cms{} when using Paxos with a leader for consensus and requesting barriers in the beginning of the protocol is $6\delta$. It is so because some messages may need to have their initial timestamp changed as a possible consequence of the optimistic assumption \emph{A2} not holding. However, a simple way to eliminate the possibility of having to change timestamps is by not considering messages whose delivery order does not follow the order of the initial timestamps.

@@

The problem with changing the original uniform validity property is that it could make the multicast primitive impracticable for some applications. Nevertheless, some other applications not only do not require messages to be delivered if they arrive out of order, but also it is better not to deliver such messages at all -- such as real-time streams of audio or video or online real-time multiplayer games.
